---
title: "Lab 03"
author: "Robert M. Trehern"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: cerulean
    highlight: haddock
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE, results='asis'}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include=FALSE}
# load necessary packages ----
library( dplyr )
library( import )
library( here )
library( knitr )
library( pander )
library( stargazer )
library( scales )

# change s.type to "html" when knitting
s.type <- "text" 
```


```{r, include=FALSE}
import::here("clean_d",
             "tidy_up_data",
             "build_year",
             "RELEVANT_FILES",
             "obtain_crosswalk",
             "create_final_metadata_file",
             "compare_dfs",
             "jplot",
             .from = here::here("labs/wk03/utilities.R"),
             .character_only = TRUE)
```


```{r, include=FALSE}
# load data as character vectors
d.2000.samp <- read.csv( here::here("data/raw/ltdb_std_2000_sample.csv"),
                         colClasses="character" )

d.2000.samp <- clean_d( d.2000.samp, start_column=5 )
```


```{r, include=FALSE}
file.name <- "ltdb_std_2000_sample.csv"
d.2000.s <- tidy_up_data( file.name )
head( d.2000.s[1:20] ) %>% pander()
file.name <- "LTDB_Std_2000_fullcount.csv"
d.2000.f <- tidy_up_data( file.name )
head( d.2000.f[1:20] ) %>% pander()
d2 <- bind_rows( d.2000.s, d.2000.f )
file.name <- "ltdb_std_2000_sample.csv"
d.2000.s <- tidy_up_data( file.name )
head( d.2000.s[1:20] ) %>% pander()
file.name <- "LTDB_Std_2000_fullcount.csv"
d.2000.f <- tidy_up_data( file.name )
head( d.2000.f[1:20] ) %>% pander()
d2 <- bind_rows( d.2000.s, d.2000.f )
```


```{r,include=FALSE}
d <- readRDS( here::here( "data/rodeo/LTDB-1990.rds" ) )
head( d ) %>% pander()
```

## Part 01 - Change in Home Values

```{r}
# load and clean data for analysis

d1 <- readRDS( here::here( "data/rodeo/LTDB-1990.rds" ) )
d2 <- readRDS( here::here( "data/rodeo/LTDB-2000.rds" ) )
md <- readRDS( here::here( "data/rodeo/LTDB-META-DATA.rds" ) )

d1 <- select( d1, - year )
d2 <- select( d2, - year )

d <- merge( d1, d2, by="tractid" )
d <- merge( d, md, by="tractid" )

d <- filter( d, urban == "urban" ) # filter rural districts
```

```{r, include = FALSE}
vars <- compare_dfs( df1=d1, df2=d2 ) # identify common variables
```

<br>

```{r}
# create dataset for analysis

d.full <- d

d <- d.full

d <- select( d, tractid, mhmval90, mhmval00, hinc90, 
             hu90, own90, rent90,  
             empclf90, clf90, unemp90, prof90,  
             dpov90, npov90,
             ag25up90, hs90, col90, 
             pop90.x, nhwht90, nhblk90, hisp90, asian90,
             cbsa, cbsaname )
d <- 
  d %>%
  mutate( p.white = 100 * nhwht90 / pop90.x,
          p.black = 100 * nhblk90 / pop90.x,
          p.hisp = 100 * hisp90 / pop90.x, 
          p.asian = 100 * asian90 / pop90.x,
          p.hs = 100 * (hs90+col90) / ag25up90,
          p.col = 100 * col90 / ag25up90,
          p.prof = 100 * prof90 / empclf90,
          p.unemp = 100 * unemp90 / clf90,
          pov.rate = 100 * npov90 / dpov90 )
```

<br>

```{r}
stargazer( d, 
           type=s.type, 
           digits=0,
           summary.stat = c("min", "p25","median","mean","p75","max") )
```

<br>

```{r}
# calculate inflation rate
# inflation rate between 1990 and 2000 = 3.1%
(1.031)^10

# adjust 1990 home values for inflation 
mhv.90 <- d$mhmval90 * 1.357021  
mhv.00 <- d$mhmval00

mhv.change <- mhv.00 - mhv.90

df <- data.frame( MedianHomeValue1990=mhv.90, 
                  MedianHomeValue2000=mhv.00, 
                  Change.90.to.00=mhv.change )

stargazer( df, 
           type=s.type, 
           digits=0, 
           summary.stat = c("min", "p25","median","mean","p75","max") )
```

<br>

```{r}
hist( mhv.change/1000, breaks=500, 
      xlim=c(-200,300), yaxt="n", xaxt="n",
      xlab="Thousand of US Dollars (adjusted to 2000)", cex.lab=1.5,
      ylab="", main="Change in Median Home Value 1990 to 2000",
      col="gray20", border="white" )

axis( side=1, at=seq( from=-200, to=300, by=100 ), 
      labels=paste0( "$", seq( from=-200, to=300, by=100 ), "k" ) )
        
mean.x <- mean( mhv.change/1000, na.rm=T )
abline( v=mean.x, col="darkorange", lwd=2, lty=2 )
text( x=200, y=1500, 
      labels=paste0( "Mean = ", dollar( round(1000*mean.x,0)) ), 
      col="darkorange", cex=1.8, pos=3 )

median.x <- median( mhv.change/1000, na.rm=T )
abline( v=median.x, col="dodgerblue", lwd=2, lty=2 )
text( x=200, y=2000, 
      labels=paste0( "Median = ", dollar( round(1000*median.x,0)) ), 
      col="dodgerblue", cex=1.8, pos=3 )
```

<br>

```{r}
layout.matrix <- matrix( c( 1,3,
                            2,3 ), 
                nrow=2, ncol=2, byrow=T )

layout( mat = layout.matrix,
        heights = c(2,2), # Heights of the two rows
        widths =  c(3,4)) # Widths of the two columns

# layout.show(3)

par( mar=c(4,0,0,2) )

hist( mhv.90/1000, breaks=50, 
      xlim=c(-200,800), yaxt="n", xaxt="n",
      xlab="", cex.lab=1,
      ylab="", main="",
      col="darkslateblue", border="white" )

axis( side=1, at=seq( from=0, to=1000, by=100 ), 
      labels=paste0( "$", seq( from=0, to=1000, by=100 ), "k" ) )

abline( v=seq(0,1000,100), lty=2, col="gray80" )

text( 550, 4000, labels="Median Home \nValue in 1990", 
      col="darkslateblue", cex=1.8 )



hist( mhv.00/1000, breaks=50, 
      xlim=c(-200,800), yaxt="n", xaxt="n",
      xlab="", cex.lab=1,
      ylab="", main="",
      col="darkslateblue", border="white" )

abline( v=seq(0,1000, 100 ), lty=2, col="gray80" )

text( 550, 3500, labels="Median Home \nValue in 2000", 
      col="darkslateblue", cex=1.8 )

axis( side=1, at=seq( from=0, to=1000, by=100 ), 
      labels=paste0( "$", seq( from=0, to=1000, by=100 ), "k" ) )


# data reduction - filter 1,000 observations

df <- data.frame( v90=mhv.90/1000, v00=mhv.00/1000 )
df <- sample_n( df, 1000 )

par( mar=c(4,5,3,2) )

jplot( df$v90, df$v00, 
       lab1="MHV in 1990", lab2="MHV in 2000",
       xlim=c(0,1000), ylim=c(0,1000),
       axes=F )

abline( a=0, b=1, lty=2, col="gray" )
axis( side=1, at=seq( from=0, to=1000, by=200 ), 
      labels=paste0( "$", seq( from=0, to=1000, by=200 ), "k" ) )
axis( side=2, at=seq( from=0, to=1000, by=200 ), 
      labels=paste0( "$", seq( from=0, to=1000, by=200 ), "k" ) )
```

<br>

```{r}
mhv.90[ mhv.90 < 10000 ] <- NA
pct.change <- mhv.change / mhv.90
summary( pct.change )
```

```{r}
sum( pct.change > 5, na.rm=T )
```

```{r}
d %>% 
  filter( pct.change > 5 ) %>% 
  head()
```

```{r}
hg <-
hist( pct.change, breaks=2000, 
      xlim=c(-1,2), yaxt="n", xaxt="n",
      xlab="", cex.main=1.5,
      ylab="", main="Growth in Home Value by Census Tract 1990 to 2000",
      col="gray40", border="white" )

axis( side=1, at=seq( from=-1, to=2, by=0.5 ), 
      labels=paste0( seq( from=-100, to=200, by=50 ), "%" ) )

ymax <- max( hg$count )
        
mean.x <- mean( pct.change, na.rm=T )
abline( v=mean.x, col="darkorange", lwd=2, lty=2 )
text( x=1, y=(0.5*ymax), 
      labels=paste0( "Mean = ", round(100*mean.x,0), "%"), 
      col="darkorange", cex=1.8, pos=4 )

median.x <- median( pct.change, na.rm=T )
abline( v=median.x, col="dodgerblue", lwd=2, lty=2 )
text( x=1, y=(0.6*ymax), 
      labels=paste0( "Median = ", round(100*median.x,0), "%"), 
      col="dodgerblue", cex=1.8, pos=4 )
```

<br>

```{r}
d$mhv.change <- mhv.change 
d$pct.change <- pct.change
d$mhv.00 <- mhv.00
d$mhv.90 <- mhv.90

d %>%
  group_by( cbsaname ) %>%
  summarize( ave.change = median( mhv.change, na.rm=T ),
             ave.change.d = dollar( round(ave.change,0) ),
             growth = 100 * median( pct.change, na.rm=T ) ) %>%
  ungroup() %>%
  arrange( - growth ) %>%
  select( - ave.change ) %>% 
  head( 25 ) %>%
  pander()
```

##### Part 01 - Questions:

##### How do changes in home value differ between the 1990-2000 period and 2000-2010?
**Between 1990-2000, the median home value fell $2,239 on average.  Between 2000-2010, the median home value increased by $36,268.**   

##### What do the authors suggest would predict fall in central city home values between 1990 and 2000?
**The authors suggest shifting demographics, primarily measured by the fraction of educated white population, would predict a decrease in central city home values.**  

<br>
<br>

----

<br>
<br>

### Part 02 - Measuring Gentrification 

##### Select Gentrification Variables

```{r}
# adjust 2000 home values for inflation 
mhv.90 <- d.full$mhmval90 * 1.031  
mhv.00 <- d.full$mhmval00

mhv.change <- mhv.00 - mhv.90

# small initial values are skewing percentages
#
# an average home value below $10k is really low -
# these must be mostly vacant lots?

mhv.90[ mhv.90 < 10000 ] <- NA
pct.change <- 100 * ( mhv.change / mhv.90 )
summary( pct.change )
##     Min.  1st Qu.   Median     Mean  3rd Qu.     Max.     NA's 
##  -96.891    5.918   25.402   33.167   49.556 6059.261      220
d.full$mhv.90 <- mhv.90
d.full$mhv.00 <- mhv.00
d.full$mhv.change <- mhv.change
d.full$pct.change <- pct.change
```

<br>

```{r}
d3 <- select( d.full, tractid, cbsa, cbsaname,
              mhv.90, mhv.00, mhv.change, pct.change,              # home value
              hinc90, hu90, own90, rent90, incpc90,                # ses
              hinc00, hu00, own00, rent00, incpc00,
              empclf90, clf90, unemp90, prof90, manuf90,           # employment
              empclf00, clf00, unemp00, prof00, manuf00,
              dfmpov90, dpov90, npov90, dfmpov00, dpov00, npov00,  # poverty
              ag25up90, hs90, col90, ag25up00, hs00, col00,        # education
              pop90.x, nhwht90, nhblk90, hisp90, asian90, ntv90,   # race
              pop00.x, nhwht00, nhblk00, hisp00, asian00, ntv00 )

d3 <- 
  d3 %>%
  mutate( 
    # 2010 variables
          p.white.90 = 100 * nhwht90 / pop90.x,
          p.black.90 = 100 * nhblk90 / pop90.x,
          p.hisp.90 = 100 * hisp90 / pop90.x, 
          p.asian.90 = 100 * asian90 / pop90.x,
          p.hs.edu.90 = 100 * (hs90+col90) / ag25up90,
          p.col.edu.90 = 100 * col90 / ag25up90,
          p.prof.90 = 100 * prof90 / empclf90,
          p.unemp.90 = 100 * unemp90 / clf90,
          pov.rate.90 = 100 * npov90 / dpov90,
          p.hinc.90 = 100 * hinc90 / hu90,
          
          # 2000 variables
          p.white.00 = 100 * nhwht00 / pop00.x,
          p.black.00 = 100 * nhblk00 / pop00.x,
          p.hisp.00 = 100 * hisp00 / pop00.x, 
          p.asian.00 = 100 * asian00 / pop00.x,
          p.hs.edu.00 = 100 * (hs00+col00) / ag25up00,
          p.col.edu.00 = 100 * col00 / ag25up00,
          p.prof.00 = 100 * prof00 / empclf00,
          p.unemp.00 = 100 * unemp00 / clf00,
          pov.rate.00 = 100 * npov00 / dpov00,
          p.hinc.00 = 100 * hinc00 / hu00)
```

<br>

```{r}
d3 <-
  d3 %>%
  group_by( cbsaname ) %>%
  mutate( metro.mhv.pct.90 = ntile( mhv.90, 100 ),
          metro.mhv.pct.00 = ntile( mhv.00, 100 ),
          metro.median.pay.90 = median( hinc90, na.rm=T ),
          metro.median.pay.00 = median( hinc00, na.rm=T ),
          metro.race.rank.90 = ntile( (100-p.white.90), 100 ) ) %>%
  ungroup() %>%
  mutate( metro.mhv.pct.change = metro.mhv.pct.00 - metro.mhv.pct.90,
          pay.change = metro.median.pay.00 - metro.median.pay.90,
          race.change = p.white.00 - p.white.90,
          mhv.change = mhv.00 - mhv.90 )
```

<br>

```{r}
d3 <-           
  d3 %>%
  select( c( "tractid", "cbsa", "cbsaname",
             "mhv.90", "mhv.00", "mhv.change","pct.change",
          "p.white.90", "p.black.90", "p.hisp.90", "p.asian.90", 
          "p.hs.edu.90", "p.col.edu.90", "p.prof.90",  "p.unemp.90", 
          "pov.rate.90", "p.hinc.90", "p.white.00", "p.black.00", "p.hisp.00", 
          "p.asian.00", "p.hs.edu.00", "p.col.edu.00", "p.prof.00", 
          "p.unemp.00", "pov.rate.00", "p.hinc.00", "metro.mhv.pct.90", 
          "metro.mhv.pct.00", "metro.median.pay.90", "metro.median.pay.00", 
          "metro.mhv.pct.change", "pay.change", "race.change",
          "metro.race.rank.90") ) 
```

<br>

```{r}
d3 <- data.frame(d3)
stargazer( d3, 
           type=s.type, 
           digits=0, 
           summary.stat = c("min", "p25","median","mean","p75","max") )
```

<br>

```{r}
# home value in lower than average home in a metro in 1990
poor.1990 <- d3$metro.mhv.pct.90 < 50  

# above average diversity for metro area
diverse.1990 <- d3$metro.race.rank.90 > 50 

# home values increased more than overall city gains 
# change in percentile rank within the metro
mhv.pct.increase <- d3$metro.mhv.pct.change > 0

# faster than average growth  
# 25% growth in value is median for the country
home.val.rise <- d3$pct.change > 25 

# proportion of whites increases by more than 3 percent 
# measured by increase in white
loss.diversity <- d3$race.change > 3 

g.flag <- poor.1990 & diverse.1990 & mhv.pct.increase & home.val.rise & loss.diversity

num.candidates <-  sum( poor.1990 & diverse.1990, na.rm=T )
num.gentrified <- sum( g.flag, na.rm=T )

num.gentrified 
```

<br>

```{r}
num.candidates
```

<br>

```{r}
num.gentrified / num.candidates
```

<br>

```{r}
mhv.90[ mhv.90 < 1000 ] <- NA
pct.change <- 100 * ( mhv.change / mhv.90 )
summary( pct.change )
```


##### Part 02 - Questions:

##### How many census tracts are candidates (start out at a low income level with high diversity)?
**Based on the variables, there were 17,560 candidates for gentrification.**

##### How many have transitioned into advanced stages of gentrification?
**Only 600 were considered gentrified from 1990 to 2000.  The gentrification rate was approximately 3.4%.**

<br>
<br>

----

<br>
<br>

### Part 03 - Spatial Patterns




